SPANG Template Library: Making SPARQL Queries FAIR

### 2020-05-23 paper.tex
v1.3

\begin{document}
%
\title{SPANG Library: Making SPARQL Queries FAIR}
%
\titlerunning{SPANG library: making SPARQL FAIR}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
\author{Hirokazu Chiba}
%
\authorrunning{H. Chiba}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
%\institute{Princeton University, Princeton NJ 08544, USA \and
%Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
%\email{lncs@springer.com}\\
%\url{http://www.springer.com/gp/computer-science/lncs} \and
%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%\email{\{abc,lncs\}@uni-heidelberg.de}}
\institute{Database Center for Life Science, Chiba 277-0871, Japan\\
\email{chiba@dbcls.rois.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 15--250 words.
How can we maximize the value of accumulated SPARQL queries? 
%SPARQL is a key component of the Semantic Web, thus the reusability of SPARQL is important.
An increasing number of datasets have been published in the Resource Description Framework (RDF) and the SPARQL standard allows interoperable framework for querying the RDF datasets; however, reuse of SPARQL queries remains limited partly due to the lack of accessibility and findability of SPARQL queries that were once written. 
%SPARQL contributes to the standardized interface to the RDF, and maximizing the reusability of RDF; however, reusability of SPARQL itself remains to be addressed. 
In this study, we focus on FAIR (findable, accessible, interoperable,
reusable) principle to improve usability of SPARQL queries.
We present the process of making SPARQL queries FAIR and demonstrate the usefulness of the result.
FAIRification of the SPARQL is a plausible way to make SPARQL interoperable and reusable.
Here, we define a SPANG library as a controleld query set with specific
extension of SPARQL so that the query set follows FAIR pinciple.
%Here we propose a SPARQL extension to construct a reusable query library to enrich the Semantic Web platform.
Our contribution is the improvement of findability and accessibility of SPARQL queries.
1) We present a metadata set for annotating a SPARQL query, which
enables to identify SPARQL query and provides useful information for the query.
2) We present a mechanism to parameterize the SPARQL query.
It makes each query reusable in different settings.
This study make SPARQL queries FAIR, and this specification leads to the construction of eco-system to reuse SPARQL queries to enrich the use of the Semantic Web platform.

@@

\end{abstract}
%
%
%
\section{Introduction}
How can we maximize the value of accumulated SPARQL queries?
Reuse of them of could be the key.
An increasing number of datasets have been published in the Resource Description Framework (RDF).
SPARQL is the standardized interface to RDF and contributing the reusability of the RDF data. 
Although the SPARQL language provides a basis for exploiting RDF databases, writing a raw SPARQL code often becomes a burden for users. 
SPARQL is the standardized query language, and many Semantic Web users write queries in SPARQL, with a great amount of efforts.
Those queries are valuable resource.
Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, the reuse of those queries are limited. 
Reusing SPARQL queries are important for efficient development of the Semantic Web.

@@

% The following section includes: overview of the SPANG project, specification of SPANG templates, implementation of the SPANG library, and use case of SPANSG library.

% \begin{figure}
% \center
% \includegraphics[width=1.0\textwidth]{spang_top.png}
% \caption{SPANG web site}
% \label{fig:spang_top}
% \end{figure}

@@

% Figure~\ref{fig:spang_top} is the top page of SPANG, which includes links to GitHub, Documentation, and SPANG library.

@@

\subsection{SPANG template specification}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustash notation.

\subsection{SPARQL template engine}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used PEG~\cite{peg} expression of SPARQL grammar originally expressed in EBNF in the SPARQL specification. The original implementation of PEG.js~\cite{pegjs} of SPARQL grammar is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON object.

@@

\subsection{SPARQL template library}
%We implemented SPARQL queries for 10 endpoints.

\subsubsection{SPANG coding rules}
We defined a coding rule for SPARQL and its extension in SPANG framework. Coding rule is important for productivity of shared coding tasks and effective maintenance. It also helps to make debug process easier; weird look suggests possible bugs. 
\begin{itemize}
    \item Header lines: query can start with header lines as comment lines. Metadata can be added at the top of the query as comment lines. Multiple lines are allowed. If multiple lines are described, they are written in sequential without blank lines. After the metadata lines, a blank line comes.
    \item Prefix: prefix declaration comes in sequential lines. Each of before and after the prefix declaration lines, a blank line exists.
    \item Cases: the SPARQL keywords should be in upper case characters. variable names comes with ? followed by lower case characters.
    \item Literal: the literals are surrounded by "".
    \item Indent: indentation depth is two spaces.
    \item Spaces: a space exists between keywords and variables. A space exists before , or ; or . .
    \item Newline: a newline exists before the keyword \texttt{WHERE} and after \texttt{\{}. A new line exist after ; or . but not after ,. After the newline following ;, an indent follows. A newline exist before query modifiers such as \texttt{ORDER BY} or \texttt{LIMIT}.
\end{itemize}
We developed a reformatter of the SPARQL code as described below.

@@

\subsection{SPANG Web site}
SPARQL queries can be executed in Web site.

%%%%%%%%%%%%%%%%%%%%

% \subsection{Simple querying with SPARQL shortcuts}

% Specific command-line options, including \texttt{-S~\textit{SUBJECT}}, \texttt{-P~\textit{PREDICATE}},\\ \texttt{-O~\textit{OBJECT}}, \texttt{-L~\textit{LIMIT}}, and other modifiers, 
% are interpreted as shortcuts for generating typical SPARQL queries. 
% A simple example is:

% \texttt{spang uniprot -S tax:511145 -a}

% \noindent where the first argument is the target SPARQL endpoint and the following arguments are SPARQL shortcuts (the SPARQL endpoint can be specified in a URL or nickname for simplicity). 
% \texttt{uniprot} is a nickname for the UniProt SPARQL endpoint \cite{UniProt} and \texttt{tax:} is a prefix for URIs of NCBI taxonomy IDs. 
% This example command line searches the target database for statements that have taxonomy ID 511145 ({\it Escherichia coli} strain K-12 MG1655) as a subject. Using the \texttt{-a} option transforms the URIs in the search result into abbreviated forms using predefined prefix declarations. For example,\\
% \texttt{\textless http://www.w3.org/2000/01/rdf-schema\#subClassOf\textgreater}\\
% is transformed into \texttt{rdfs:subClassOf}.
% The result is output to the standard output in the form of tab-separated values by default, as it is suitable for processing by line-oriented Unix programs.
% In addition, combination of subject, predicate, and object is possible according to the following.

% \texttt{spang uniprot -S tax:511145 -P up:otherName}

% \noindent where predicate {\tt up:otherName} is specified to further confine the results to other names. 
% For the full list of available command-line options, simply type the command {\tt spang}.

% \subsection{Combinatorial execution of multiple queries}

% SPANG combines multiple queries through a Unix pipe. 
% Combining a {\tt spang} command in shortcut mode and another one in template mode is also possible.
% An example of such a combinatorial query is shown in Fig.~\ref{fig:01}.
% The first {\tt spang} command is in template mode, which is the same as that presented in the Section 2.3. 
% This command searches the MBGD database for orthologs of a protein {\tt K9Z723}. 
% The obtained list of proteins are used in the second query, which searches the UniProt database for annotations of the given list of proteins. 
% The option \texttt{-S 1} specifies the values from the first column of the standard input as subject.
% This combinatorial query enables integrative use of two databases distributed across the Web. 
% Here, the output of the first command can be further modified by altering the second command.
% For example:
% \begin{quoting}
% \texttt{spang mbgd mbgdl:get\_ortholog K9Z723 | spang uniprot uniprot\_xref PDB}
% \vspace{1pt}
% \end{quoting}
% where {\tt uniprot\_xref} is a SPARQL template (see supplementary data for the code), which retrieves cross-references from the UniProt IDs given in the standard input to the database specified as the parameter (in this example, {\tt PDB}). This example command line searches for entries in the Protein Data Bank (PDB) \citep{Berman} among orthologs of {\tt K9Z723}.

\subsection{SPANG portal web site}
SPANG portal web site is developed using Ruby on Rails. And it can be run on Docker. The SPANG query library is published on GitHub. Thus, users can run a portal site by their own.

% \subsection{SPANG command: simple querying with shortcuts}

% Specific command-line options, including \texttt{-S~\textit{SUBJECT}}, \texttt{-P~\textit{PREDICATE}}, \texttt{-O~\textit{OBJECT}}, \texttt{-L~\textit{LIMIT}}, and other modifiers, 
% are interpreted as shortcuts for generating typical SPARQL queries. 
% A simple example is:

% \texttt{spang uniprot -S tax:511145 -a}

% \noindent where the first argument is the target SPARQL endpoint and the following arguments are SPARQL shortcuts (the SPARQL endpoint can be specified in a URL or nickname for simplicity). 
% \texttt{uniprot} is a nickname for the UniProt SPARQL endpoint \citep{UniProt} and \texttt{tax:} is a prefix for URIs of NCBI taxonomy IDs. 
% This example command line searches the target database for statements that have taxonomy ID 511145 ({\it Escherichia coli} strain K-12 MG1655) as a subject. Using the \texttt{-a} option transforms the URIs in the
%  search result into abbreviated forms using predefined prefix declarations. For example,\\
% \texttt{\textless http://www.w3.org/2000/01/rdf-schema\#subClassOf\textgreater} 
% is transformed into \texttt{rdfs:subClassOf
% }.
% The result is output to the standard output in the form of tab-separated values by default, as it is suitable for processing by line-oriented Unix programs.
% In addition, combination of subject, predicate, and object is possible according to the following.

% \texttt{spang uniprot -S tax:511145 -P up:otherName}

% \noindent where predicate {\tt up:otherName} is specified to further confine the results to other names. 
% For the full list of available command-line options, simply type the command {\tt spang}.
% \vspace*{-5pt}

@@

% \subsection{A Subsection Sample}
% Please note that the first paragraph of a section or subsection is
% not indented. The first paragraph that follows a table, figure,
% equation etc. does not need an indent, either.

% Subsequent paragraphs, however, are indented.

% \subsubsection{Sample Heading (Third Level)} Only two levels of
% headings should be numbered. Lower level headings remain unnumbered;
% they are formatted as run-in headings.

% \paragraph{Sample Heading (Fourth Level)}
% The contribution should contain no more than four levels of
% headings. Table~\ref{tab1} gives a summary of all heading levels.

% \begin{table}
% \caption{Table captions should be placed above the
% tables.}\label{tab1}
% \begin{tabular}{|l|l|l|}
% \hline
% Heading level &  Example & Font size and style\\
% \hline
% Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
% 1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
% 2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
% 3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
% 4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
% \hline
% \end{tabular}
% \end{table}

% \noindent Displayed equations are centered and set on a separate
% line.
% \begin{equation}
% x + y = z
% \end{equation}
% Please try to avoid rasterized images for line-art diagrams and
% schemas. Whenever possible, use vector graphics instead (see
% Fig.~\ref{fig1}).

% \begin{figure}
% \includegraphics[width=\textwidth]{fig1.eps}
% \caption{A figure caption is always placed below the illustration.
% Please note that short captions are centered, while long ones are
% justified by the macro package automatically.} \label{fig1}
% \end{figure}

% \begin{theorem}
% This is a sample theorem. The run-in heading is set in bold, while
% the following text appears in italics. Definitions, lemmas,
% propositions, and corollaries are styled the same way.
% \end{theorem}
%
% the environments 'definition', 'lemma', 'proposition', 'corollary',
% 'remark', and 'example' are defined in the LLNCS documentclass as well.
%
% \begin{proof}
% Proofs, examples, and remarks have the initial word in italics,
% while the following text appears in normal font.
% \end{proof}
% For citations of references, we prefer the use of square brackets
% and consecutive numbers. Citations using labels or the author/year
% convention are also acceptable. The following bibliography provides
% a sample reference list with entries for journal
% articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
% book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
% and a homepage~\cite{ref_url1}. Multiple citations are grouped
% \cite{ref_article1,ref_lncs1,ref_book1},
% \cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\begin{thebibliography}{8}

% \bibitem{Rietveld}
% Rietveld,L. and Hoekstra,R.: YASGUI: not just another SPARQL client,
% In: The Semantic Web: ESWC 2013 Satellite Events, 78-86. Springer (2013)

% \bibitem{Yamaguchi}
% Yamaguchi,A. \textit{et~al}.:
% An intelligent SPARQL query builder for exploration of various life-science databases.
% In: The 3rd International Workshop on Intelligent Exploration of Semantic Data (IESD 2014) (2014)

@@

% \bibitem{ref_article1}
% Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

% \bibitem{ref_lncs1}
% Author, F., Author, S.: Title of a proceedings paper. In: Editor,
% F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
% Springer, Heidelberg (2016). \doi{10.10007/1234567890}

% \bibitem{ref_book1}
% Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
% Location (1999)

% \bibitem{ref_proc1}
% Author, A.-B.: Contribution title. In: 9th International Proceedings
% on Proceedings, pp. 1--2. Publisher, Location (2010)

% \bibitem{ref_url1}
% LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
% Oct 2017
\end{thebibliography}
\end{document}

>>>>

\begin{document}

\title{SPANG Template Library: Making SPARQL Queries FAIR}
\titlerunning{SPANG: Making SPARQL Queries FAIR}
\author{Hirokazu Chiba}
\authorrunning{H. Chiba}
\institute{Database Center for Life Science, Chiba 277-0871, Japan\\
\email{chiba@dbcls.rois.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 15--250 words.
How can we maximize the value of accumulated SPARQL queries? 
%SPARQL is a key component of the Semantic Web, thus the reusability of SPARQL is important.
An increasing number of datasets have been published in the Resource Description Framework (RDF) and the SPARQL standard allows interoperable framework for querying the RDF datasets; however, reuse of SPARQL queries remains limited partly due to the lack of accessibility and findability of SPARQL queries that were once written. 
%SPARQL contributes to the standardized interface to the RDF, and maximizing the reusability of RDF; however, reusability of SPARQL itself remains to be addressed. 
In this study, we focus on FAIR (findable, accessible, interoperable,
reusable) principle to improve usability of SPARQL queries.
We present the process of making SPARQL queries FAIR and demonstrate the usefulness of the result.
FAIRification of the SPARQL is a plausible way to make SPARQL interoperable and reusable.
Here, we define a SPANG library as a controleld query set with specific
extension of SPARQL so that the query set follows FAIR pinciple.
%Here we propose a SPARQL extension to construct a reusable query library to enrich the Semantic Web platform.
Our contribution is the improvement of findability and accessibility of SPARQL queries.
1) We present a metadata set for annotating a SPARQL query, which
enables to identify SPARQL query and provides useful information for the query.
2) We present a mechanism to parameterize the SPARQL query.
It makes each query reusable in different settings.
This study make SPARQL queries FAIR, and this specification leads to the construction of eco-system to reuse SPARQL queries to enrich the use of the Semantic Web platform.

@@

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%

How can we maximize the value of accumulated SPARQL queries?
Reuse of them is the key.
An increasing number of datasets have been published in the Resource Description Framework (RDF).
SPARQL is the standardized interface to RDF and contributing the reusability of the RDF data. 
Although the SPARQL language provides a basis for exploiting RDF databases, writing a raw SPARQL code often becomes a burden for users. 
SPARQL is the standardized query language, and many Semantic Web users write queries in SPARQL, with a great amount of efforts.
Those queries are valuable resource.
Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, the reuse of those queries are limited. 
Reusing SPARQL queries are important for efficient development of the Semantic Web.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
%%%%%%%%%%%%%%%%%%%%%%%%

We overview the FAIR principles, then describe how to apply the principle to SPARQL.
We demonstarate how to use the SPANG templete library.
Then, we describe the details of SPANG template specifications.

%The following section includes: overview of the SPANG project, specification of SPANG templates, implementation of the SPANG library, and use case of SPANSG library.

Figure~\ref{fig:spang_top} is the top page of SPANG, which includes links to GitHub, Documentation, and SPANG library.

\begin{figure}
\center
\includegraphics[width=1.0\textwidth]{spang_top.png}
\caption{SPANG web site}
\label{fig:spang_top}
\end{figure}

@@

@@

\subsection{Template specifications}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustash notation.

\subsection{Template engine}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used PEG~\cite{peg} expression of SPARQL grammar originally expressed in EBNF in the SPARQL specification. The original implementation of PEG.js~\cite{pegjs} of SPARQL grammar is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON object.

@@

\subsection{Coding rules}
We defined a coding rule for SPARQL and its extension in SPANG framework. Coding rule is important for productivity of shared coding tasks and effective maintenance. It also helps to make debug process easier; weird look suggests possible bugs. 
\begin{itemize}
    \item Header lines: query can start with header lines as comment lines. Metadata can be added at the top of the query as comment lines. Multiple lines are allowed. If multiple lines are described, they are written in sequential without blank lines. After the metadata lines, a blank line comes.
    \item Prefix: prefix declaration comes in sequential lines. Each of before and after the prefix declaration lines, a blank line exists.
    \item Cases: the SPARQL keywords should be in upper case characters. variable names comes with ? followed by lower case characters.
    \item Literal: the literals are surrounded by "".
    \item Indent: indentation depth is two spaces.
    \item Spaces: a space exists between keywords and variables. A space exists before , or ; or . .
    \item Newline: a newline exists before the keyword \texttt{WHERE} and after \texttt{\{}. A new line exist after ; or . but not after ,. After the newline following ;, an indent follows. A newline exist before query modifiers such as \texttt{ORDER BY} or \texttt{LIMIT}.
\end{itemize}
We developed a reformatter of the SPARQL code as described below.

\subsection{SPANG server}
SPARQL queries can be executed in Web site.
SPANG server is developed using Ruby on Rails. And it can be run on Docker. The SPANG query library is published on GitHub. Thus, users can run a portal site by their own.

@@

@@

\begin{thebibliography}{8}

@@

\end{thebibliography}
\end{document}
